---
title: "Untitled"
author: "Matthew Upson"
date: "31/10/2014"
output:
  html_document:
    theme: spacelab
  pdf_document: default
---
### Data Description:

* Trip Duration (seconds)
* Start Time and Date
* Stop Time and Date
* Start Station Name
* End Station Name
* Station ID
* Station Lat/Long
* Bike ID
* User Type (Customer = 24-hour pass or 7-day pass user; Subscriber = Annual Member)
* Gender (Zero=unknown; 1=male; 2=female)
* Year of Birth

```{r load_libraries,include=TRUE,echo=TRUE,message=FALSE,warning=FALSE,include=TRUE}
require(plyr)
require(dplyr)
require(lubridate)
#require(testthat)
#require(reshape2)
require(ggplot2)
require(ggmap)
#require(randomForest)
#require(xtable)
```


```{r load_stns,cache=TRUE,echo=TRUE,warning=FALSE,message=FALSE,include=TRUE}

nybikes_db <- src_sqlite(
  "~/Dropbox/R/Citibike/nybikes.db"
  )


stns_sql <- # convert db table to R dataframe
  tbl( # load a database table
    nybikes_db,
    sql("select * from stns")
    )


stns <- collect(stns_sql)

```

### Let's start by creating some maps:

```{r ggmap12,cache=TRUE,echo=TRUE,warning=FALSE,message=FALSE,include=TRUE}

# calculate centre of all points
centre <- c(
  lon = mean(stns$lon),
  lat = mean(stns$lat)
  )


nymap_12 <- get_map(
     centre, 
     scale = 2,
     zoom = 12,
     source = "google"
)

nymap_13 <- get_map(
     centre, 
     scale = 2,
     zoom = 13,
     source = "google"
)

ggmap(
     nymap_12, 
     extent = 'device',
     legend="bottomright"
) + 
  geom_point(
    data = stns,
    aes(
      x = lon,
      y = lat
      )
    )
```

```{r ggmap13,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}

ggmap(
     nymap_13, 
     extent = 'device',
     legend="bottomright"
) + 
  geom_point(
    data = stns,
    aes(
      x = lon,
      y = lat
      )
    )

```


### Initial exploration

Ok so now that we know where the stations are...let's do something with it! Let's start some of the very simple questions...

```{r load_nybikes,cache=FALSE,echo=TRUE,message=FALSE,warning=FALSE,include=TRUE}

load("bikes.RData")

# nybikes_db <- src_sqlite(
#   "~/Dropbox/R/Citibike/nybikes.db"
#   )
# 
# bikes_sql <- tbl( # load a database table
#     nybikes_db,
#     sql("select * from nybikes")
#     )
# 
# bikes <- collect(bikes_sql)
# 
# bikes <- bikes %>%
#   dplyr::mutate(
#     starttime = ymd_hms(starttime),
#     stoptime = ymd_hms(stoptime),
#     start_hash = factor(start_hash),
#     end_hash = factor(end_hash),
#     bikeid = factor(bikeid),
#     usertype = factor(usertype),
#     gender = factor(gender),
#     birth_year = ifelse(birth_year == "\\N",NA,birth_year),
#     gender = mapvalues(
#       gender, 
#       from = c(0,1,2), 
#       to = c(NA,"M","F"))
#     )

```


```{r,cache=TRUE}

bla <- bikes %>% 
  dplyr::filter(
    !is.na(gender)
    ) %>%
  dplyr::group_by(
    we = ifelse(
      wday(stoptime) %in% c(1,7),
      "Weekend",
      "Weekday"
      ),    
    year = year(stoptime),
    month = month(stoptime),
    day = day(stoptime),
    hash = as.character(end_hash),
    gender
    ) %>%
  dplyr:: summarise(
    obs = n()
    ) %>% 
  dplyr::group_by(
    gender,
    we,
    hash,
    add = FALSE
    ) %>%
  dplyr:: summarise(
    nobs = length(obs),
    obs = mean(obs)
    ) %>% 
  dplyr::left_join(
    stns
    )

cuts <- (kmeans(bla$obs,2))$centers %>% c(.,0,max(bla$obs)) %>% sort(.) %>% round(.)

bla <- bla %>%
  dplyr::mutate(
    obs_class = cut(obs,cuts)
    )

```

```{r,cache=TRUE,fig.width=10,fig.height=10}


ggmap(
  nymap_13, 
  extent = 'device',
  legend="right"
  )+
  geom_point(
    data = bla,
    aes(
      x = lon,
      y = lat,
      colour = obs_class
      ),
    size = 3,
    alpha = 0.8
    )+
  scale_colour_brewer(
    "Bike arrivals\nper day",
    palette = "Set1"
      )+facet_grid(
    gender~we
    )

```
