---
title: "Untitled"
author: "Matthew Upson"
date: "31/10/2014"
output:
  html_document:
    theme: spacelab
  pdf_document: default
---
### Data Description:

* Trip Duration (seconds)
* Start Time and Date
* Stop Time and Date
* Start Station Name
* End Station Name
* Station ID
* Station Lat/Long
* Bike ID
* User Type (Customer = 24-hour pass or 7-day pass user; Subscriber = Annual Member)
* Gender (Zero=unknown; 1=male; 2=female)
* Year of Birth

```{r,include=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
require(plyr)
require(dplyr)
require(lubridate)
#require(testthat)
#require(reshape2)
require(ggplot2)
require(ggmap)
#require(randomForest)
#require(xtable)
```

## Some interesting questions

* Are there differences in the usage pattern of men and women?
* Are there differences in the usage pattern of people of different ages?
* Do subscribers and customers use the bikes in different ways?

```{r,cache=TRUE,echo=FALSE,warning=FALSE,message=FALSE}

nybikes_db <- src_sqlite(
  "nybikes.db"
  )


stns_sql <- # convert db table to R dataframe
  tbl( # load a database table
    nybikes_db,
    sql("select * from stns")
    )


stns <- collect(stns_sql)

```

### Let's start by creating some maps:

```{r,cache=TRUE,echo=FALSE,warning=FALSE,message=FALSE}

# calculate centre of all points
centre <- c(
  lon = mean(stns$stn_long),
  lat = mean(stns$stn_lat)
  )


nymap_12 <- get_map(
     centre, 
     scale = 2,
     zoom = 12,
     source = "google"
)

nymap_13 <- get_map(
     centre, 
     scale = 2,
     zoom = 13,
     source = "google"
)

ggmap(
     nymap_12, 
     extent = 'device',
     legend="bottomright"
) + 
  geom_point(
    data = stns,
    aes(
      x = stn_long,
      y = stn_lat
      )
    )

ggmap(
     nymap_13, 
     extent = 'device',
     legend="bottomright"
) + 
  geom_point(
    data = stns,
    aes(
      x = stn_long,
      y = stn_lat
      )
    )

```

### Initial exploration

Ok so now that we know where the stations are...let's do something with it! Let's start some of the very simple questions...

```{r,cache=TRUE,echo=FALSE,message=FALSE,warning=FALSE}

nybikes_db <- src_sqlite(
  "nybikes.db"
  )

bikes_sql <- tbl( # load a database table
    nybikes_db,
    sql("select * from nybikes")
    )

bikes <- collect(bikes_sql)

bikes <- bikes %>%
  dplyr::mutate(
    starttime = ymd_hms(starttime),
    stoptime = ymd_hms(stoptime),
    start_stn_id = factor(start_stn_id),
    end_stn_id = factor(end_stn_id),
    bikeid = factor(bikeid),
    usertype = factor(usertype),
    usertype = mapvalues(
      usertype, 
      from = levels(usertype), 
      to = c("C","S")
      ),
    birth_year = ifelse("\\N",NA,birth_year),
    gender = factor(gender),
    gender = mapvalues(
      gender, 
      from = c(0,1,2), 
      to = c(NA,"M","F"))
    )

```





```{r,cache=FALSE}


ggplot(
  bikes %>%
    dplyr::sample_n(
      100000,
      replace = FALSE
      ) %>%
    dplyr::group_by(
      # Round date values to the nearest first day of the month    
      year = ordered(year(starttime)),
      month = ordered(month(starttime))
      ) %>%
    dplyr::summarise(
      n = length(tripduration)
      ),
  aes(
    x = month,
    y = n,
    colour = year
    )
  )+
  geom_path()


```

Starting with some simple questions:

* Do women spend longer on bikes than men?
* Are there differences in the number of subscribers among men and women?
* Do women use bikes at difference hours than men?
* What about on different days?

```{r,cache=TRUE}

tapply(
  bikes$tripduration,
  bikes$gender,
  median
  )

# Ok so it looks like men spend longer on the bikes, by median
# But what proportion of users are women?

tapply(
  bikes$tripduration,
  bikes$gender,
  function(x) round(length(x)/nrow(bikes),2)
  )


#...let's plot this out
# This is likely to be quite slow, so lets take a random subsample of say 100000 
# journeys and use this to compute our density estimates


ggplot(
  data = dplyr::sample_n(
    bikes,
    size = 100000,
    replace = FALSE
    ),
  aes(
    #y = tripduration,
    x = tripduration,
    colour = gender,
    fill = gender
    )
  ) + 
  geom_density(
    alpha = 0.3
    ) +
  scale_x_continuous(
    limits = c(0,3000)
    ) +
  xlab("Trip duration (s)")

# Ok so what about different days of the week?

ggplot(
  data = bikes %>% 
    dplyr::sample_n(
      size = 100000,
      replace = FALSE
      ) %>%
    dplyr::mutate(
      wday = wday(stoptime),
      we = ifelse(
        (wday %in% c(1,7)),
        "Weekend","Weekday"
        )
      ),
  aes(
    #y = tripduration,
    x = tripduration,
    colour = gender,
    fill = gender
    )
  ) + 
  geom_density(
    alpha = 0.3
    ) +
  scale_x_continuous(
    limits = c(0,3000)
    ) +
  xlab("Trip duration (s)") +
  facet_wrap(
    ~we
    )


## Ok so far so good...women use the bikes more at weekends and weekdays...

ggplot(
  data = bikes %>% 
    dplyr::sample_n(
      size = 100000,
      replace = FALSE
      ) %>%
    dplyr::mutate(
      wday = wday(stoptime),
      we = ifelse(
        (wday %in% c(1,7)),
        "Weekend","Weekday"
        )
      ) %>%
    dplyr::group_by(
      gender,we
      ) %>%
    dplyr::summarise(
      prop = round(length(we)/100000,2)
      ),
  aes(
    #y = tripduration,
    y = prop,
    x = gender,
    colour = gender,
    fill = gender
    )
  ) + 
  geom_bar(
    alpha = 0.3,
    stat = "identity"
    ) +
  facet_wrap(
    ~we
    ) + 
  ylab("Proportion of total (sampled) journeys")


```

## New questions

* There are so many few female journeys than male journeys - but women tend to spend longer on the bikes? Where are they going? Does this mean that they are NOT commuting as much?
* Customer, as opposed to subscribers tend to have a longer journey duration. Who are these people? And where are they going? Can trends in when they make their journeys, and where they are going elucidate who these people are - are they tourists for instance?