---
title: "Untitled"
author: "Matthew Upson"
date: "31/10/2014"
output:
  html_document:
    theme: spacelab
  pdf_document: default
---
### Data Description:

* Trip Duration (seconds)
* Start Time and Date
* Stop Time and Date
* Start Station Name
* End Station Name
* Station ID
* Station Lat/Long
* Bike ID
* User Type (Customer = 24-hour pass or 7-day pass user; Subscriber = Annual Member)
* Gender (Zero=unknown; 1=male; 2=female)
* Year of Birth

```{r load_libraries,include=TRUE,echo=TRUE,message=FALSE,warning=FALSE,include=TRUE}
require(plyr)
require(dplyr)
require(lubridate)
#require(testthat)
#require(reshape2)
require(ggplot2)
require(ggmap)
#require(randomForest)
#require(xtable)
```


```{r load_stns,cache=TRUE,echo=TRUE,warning=FALSE,message=FALSE,include=TRUE}

nybikes_db <- src_sqlite(
  "~/Dropbox/R/Citibike/nybikes.db"
  )


stns_sql <- # convert db table to R dataframe
  tbl( # load a database table
    nybikes_db,
    sql("select * from stns")
    )


stns <- collect(stns_sql)

```


### Initial exploration

Ok so now that we know where the stations are...let's do something with it! Let's start some of the very simple questions...

```{r load_nybikes,cache=FALSE,echo=TRUE,message=FALSE,warning=FALSE,include=TRUE}

# It is possible to call all the data from teh sql database, but much quicker to run it from an RData file

load("bikes.RData")

# nybikes_db <- src_sqlite(
#   "~/Dropbox/R/Citibike/nybikes.db"
#   )
# 
# bikes_sql <- tbl( # load a database table
#     nybikes_db,
#     sql('select * from nybikes')
#     )
# 
# bikes <- collect(bikes_sql)
# 
# bikes <- bikes %>%
#   dplyr::mutate(
#     starttime = ymd_hms(starttime),
#     stoptime = ymd_hms(stoptime),
#     start_hash = factor(start_hash),
#     end_hash = factor(end_hash),
#     bikeid = factor(bikeid),
#     usertype = factor(usertype),
#     gender = factor(gender),
#     birth_year = ifelse(
#       birth_year == "\\N",
#       NA,
#       birth_year
#       ),
#     gender = mapvalues(
#       gender, 
#       from = c(0,1,2), 
#       to = c(NA,"M","F"))
#     )

```

### Some summary information...

* The first recorded bike trip was on `r min(bikes$starttime)`, the last recorded trip was `r max(bikes$starttime)`. 
* There are `r length(unique(bikes$bikeid))` individual bikes and `r length(unique(stns$hash))` base.
* 

```{r no_of_journeys_boxplot,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}

ggplot(
  bikes %>% 
    dplyr::group_by(
      bikeid,
      year = year(starttime),
      month = month(starttime) 
      ) %>%
    dplyr::summarise(
      n = length(bikeid)
      ),
  aes(
    y = n,
    x = ordered(month),
    fill = factor(year),
    color = factor(year)
    )
  )+
  geom_boxplot(
    alpha = 0.3,
    outlier.size = 0
    )+
  facet_wrap(~year)+ 
  ylab("Number of journeys per bike")+
  xlab("Month")
  


```

## Some initial questions questions

* Are there differences in the usage pattern of men and women?
* Are there differences in the usage pattern of people of different ages?
* Do subscribers and customers use the bikes in different ways?

```{r number_of_journeys,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}

ggplot(
  bikes %>%
    dplyr::group_by(
      # Round date values to the nearest first day of the month    
      year = ordered(year(starttime)),
      month = ordered(month(starttime))
      ) %>%
    dplyr::summarise(
      n = length(tripduration)
      ),
  aes(
    x = month,
    y = n,
    colour = year,
    group = year
    )
  )+
geom_line()+
  ylab(
    "Number of journeys"
    )+
  scale_y_continuous(
    labels = sprintf("%.0f",seq(20000,100000,20000))
    )
       

```

Starting with some simple questions:

* Do women spend longer on bikes than men?
* Are there differences in the number of subscribers among men and women?
* Do women use bikes at difference hours than men?
* What about on different days?

```{r eval=TRUE,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}

tapply(
  bikes$tripduration,
  bikes$gender,
  median
  )

# Ok so it looks like men spend longer on the bikes, by median
# But what proportion of users are women?

tapply(
  bikes$tripduration,
  bikes$gender,
  function(x) round(length(x)/nrow(bikes),2)
  )


#...let's plot this out
# This is likely to be quite slow, so lets take a random subsample of say 100000 
# journeys and use this to compute our density estimates
```

```{r trip_duration_density,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}

ggplot(
  data = bikes,
  aes(
    #y = tripduration,
    x = tripduration,
    colour = gender,
    fill = gender,
    )
  ) + 
  geom_density(
    alpha = 0.3
    ) +
  scale_x_continuous(
    limits = c(0,3000)
    ) +
  xlab("Trip duration (s)")

# Ok so what about different days of the week?
```

```{r trip_duration_density_we,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}
ggplot(
  data = bikes %>%
    dplyr::mutate(
      wday = wday(stoptime),
      we = ifelse(
        (wday %in% c(1,7)),
        "Weekend","Weekday"
        )
      ),
  aes(
    #y = tripduration,
    x = tripduration,
    colour = gender,
    fill = gender
    )
  ) + 
  geom_density(
    alpha = 0.3
    ) +
  scale_x_continuous(
    limits = c(0,3000)
    ) +
  xlab("Trip duration (s)") +
  facet_wrap(
    ~we
    )
```

```{r tripdensity_we_age,cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE,fig.width=14,fig.height=10}


ggplot(
  data = bikes %>%
    dplyr::mutate(
      wday = wday(stoptime),
      we = ifelse(
        (wday %in% c(1,7)),
        "Weekend","Weekday"
        ),
      age = 2014 - as.numeric(birth_year),
      age_group = cut(age,c(10,20,30,40,50,60,100))
      ) %>%
    dplyr::filter(
      !is.na(age_group),
      !is.na(gender)
      ),
  aes(
    #y = tripduration,
    x = tripduration,
    colour = gender,
    fill = gender
    )
  ) + 
  geom_density(
    aes(y= ..count..),
    alpha = 0.3
    ) +
  scale_x_continuous(
    limits = c(0,3000)
    ) +
  xlab("Trip duration (s)") +
  facet_grid(
    we~age_group
    )+
  ylab("Count")

```


```{r cache=TRUE,message=FALSE,warning=FALSE,include=TRUE,echo=TRUE}
## Ok so far so good...women use the bikes more at weekends and weekdays...

ggplot(
  data = bikes %>% 
    dplyr::sample_n(
      size = 100000,
      replace = FALSE
      ) %>%
    dplyr::mutate(
      wday = wday(stoptime),
      we = ifelse(
        (wday %in% c(1,7)),
        "Weekend","Weekday"
        )
      ) %>%
    dplyr::group_by(
      gender,we
      ) %>%
    dplyr::summarise(
      prop = round(length(we)/100000,2)
      ),
  aes(
    #y = tripduration,
    y = prop,
    x = gender,
    colour = gender,
    fill = gender
    )
  ) + 
  geom_bar(
    alpha = 0.3,
    stat = "identity"
    ) +
  facet_wrap(
    ~we
    ) + 
  ylab("Proportion of total (sampled) journeys")


```

## New questions

* There are so many few female journeys than male journeys - but women tend to spend longer on the bikes? Where are they going? Does this mean that they are NOT commuting as much?
* Customer, as opposed to subscribers tend to have a longer journey duration. Who are these people? And where are they going? Can trends in when they make their journeys, and where they are going elucidate who these people are - are they tourists for instance?